\input{tables/projects}
\input{tables/workshops}

\section{Workshop Experience and Terminology}
\label{sec:experience}

To develop the \workshop framework proposed in this paper, we gathered researchers who used workshops on 3 continents over the past 10 years. Our collective experience includes 17 workshops in 10 contexts: 15 workshops in 8 applied collaborations, summarized in Table~\ref{tab:projects} and Table~\ref{tab:workshops}; and 2 participatory workshops at IEEE VIS that focused on creating visualizations for domain specialists~\cite{Rogers2016,Rogers2017}.

The ways in which we use workshops have evolved over 10 years. In three of our projects, we used a series of workshops to explore opportunities, develop and iterate on prototypes, and evaluate the resulting visualizations in collaborations with cartographers~\cite{Dykes2010}, energy analysts~\cite{Goodwin2013}, and defense analysts~\cite{Walker2013}. In three additional projects, we used a single workshop to jump-start applied collaborations with neuroscientists~\cite{Kerzner2017}, constraint programmers~\cite{Goodwin2016}, and psychiatrists~\cite{Nobre2017}.  Recently, we used two workshops to explore opportunities for funded collaboration with genealogists~\cite{Kerzner2017:utdb} and biologists~\cite{Lisle2017}.

In our meta-analysis, we focused on the workshops used in the early stages of applied work or as the first in a series of workshops. To describe these workshops, we developed the term \workshops because they aim to deliberately and explicitly foster creativity while exploring opportunities for applied visualization collaborations.

Focused on \workshops, our experience includes the eight workshops in Table~\ref{tab:workshops}. Since we analyzed more data than appeared in any resulting publications, including artifacts and experiential knowledge, we refer to workshops and their projects by identifiers, e.g., [\ref{pro:edina}] refers to our collaboration with cartographers. In projects where we used more than one workshop [\ref{pro:edina} -- \ref{pro:htva}], the identifier corresponds to the {\it first} workshop in the series, unless otherwise specified. 
% Requirements workshops typically focus on eliciting opportunities for visualization software from collaborators. In general, they support the {\it understand} and {\it ideate} design activities of the design activity framework~\cite{McKenna2014} or fulfill the {\it winnow}, {\it cast}, and {\it discover} stages of the design study methodology's nine-stage framework~\cite{Sedlmair2012}.

To describe our experience, we developed terminology for the role of researchers involved in each project. The {\bf primary researcher} is responsible for deciding to use a \workshop, executing it, and integrating its results into a collaboration. Alternatively, {\bf supporting researchers} provide guidance and support to the primary researcher. We have been involved with projects as both primary and supporting researchers (see Table~\ref{tab:projects}).  

We also adopt terminology to describe \workshops. Workshops are composed of {\bf methods} --- specific, repeatable and modular activities~\cite{Crotty1998}. The methods are designed around a {\bf theme} that identifies the workshop's central topic or purpose~\cite{Brooks-Harris1999}. The {\bf facilitators} plan and guide the workshop, and the {\bf participants} carry out the workshop methods. Typically the facilitators are visualization researchers and participants are domain collaborators, but, visualization researchers can participate [\ref{pro:edina},~\ref{pro:htva}], and collaborators can facilitate [\ref{pro:cp},~\ref{pro:arbor}]. We adopted and refined this vocabulary during our reflective analysis.